---
title: "Take-home Exercise 1"
description: |
  In this Take-home Exercise 1, I will explore...
author:
  - name: Clarence Tay 
    url: https://www.linkedin.com/in/clarencetay/
    affiliation: Singapore Management University - School of Computing and Information Systems
    affiliation_url: https://scis.smu.edu.sg/
date: "`r Sys.Date()`"
output: distill::distill_article
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Overview

In this take-home exercise, blaa blab blac blad blae blaf blag blah blai blaj

# Getting Started

Before we get started, it is important for us to ensure that the required R packages have been installed. If yes, we will load the R packages. If they have yet to be installed, we will install the R packages and load them onto R environment.

The chunk code below will do the trick.

```{r}
packages = c('tidyverse')

for(p in packages){
  if(!require(p, character.only = T)){
    install.packages(p)
    }
  library(p, character.only = T)
  }
```


# Importing Data

The code chunk below imports *Participants.csv* from the data folder, into R by using [`read_csv()`](https://readr.tidyverse.org/reference/read_delim.html) of [**readr**](https://readr.tidyverse.org/) and save it as an tibble dataframe called *participants_data*.

```{r}
participants_data = read_csv("data/Participants.csv")
```

# A Simple Bar Chart

The code chunk below plots a bar chart by using geom_bar() of ggplot2.

```{r}
ggplot(data = participants_data,
       aes(x=educationLevel)) +
  geom_bar()
  
```











