---
title: "Take-home Exercise 1"
description: |
  In this Take-home Exercise 1, I will be exploring and revealing the demographic of the city of Engagement, Ohio USA by using appropriate static statistical graphics methods. 
author:
  - name: Clarence Tay 
    url: https://www.linkedin.com/in/clarencetay/
    affiliation: Singapore Management University - School of Computing and Information Systems
    affiliation_url: https://scis.smu.edu.sg/
date: "`r Sys.Date()`"
output: distill::distill_article
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      eval = TRUE, 
                      warning = FALSE, 
                      message = FALSE)
```

# Overview

In this take-home exercise, I will be exploring and revealing the demographic of the city of Engagement, Ohio USA by using appropriate static statistical graphics methods. The data will be processed by using appropriate tidyverse family of packages and the statistical graphics will be done with ggplot2 and its extensions. Datasets used will be taken from the VAST challenge 2022.

# Getting Started

Before we get started, it is important for us to ensure that the required R packages have been installed. If yes, we will load the R packages. If they have yet to be installed, we will install the R packages and load them onto R environment.

The chunk code below will do the trick.

```{r echo = FALSE}
packages = c('tidyverse')

for(p in packages){
  if(!require(p, character.only = T)){
    install.packages(p)
    }
  library(p, character.only = T)
  }
```


# Importing Data

The code chunk below imports *Participants.csv* from the data folder, into R by using [`read_csv()`](https://readr.tidyverse.org/reference/read_delim.html) of [**readr**](https://readr.tidyverse.org/) and save it as an tibble dataframe called *participants_data*.

```{r}
participants_data = read_csv("data/Participants.csv")
```

```{r}
colnames(participants_data)
count(participants_data)
count(participants_data,educationLevel)
count(participants_data,householdSize)
count(participants_data,interestGroup)
min(participants_data$age)
max(participants_data$age)
min(participants_data$joviality)
max(participants_data$joviality)
```

# A Simple Bar Chart

The code chunk below plots a bar chart by using geom_bar() of ggplot2.

```{r}
ggplot(data = participants_data,
       aes(x=age, fill= educationLevel)) +
  geom_histogram(bins=20, color='grey30')
  
```

```{r}
ggplot(data = participants_data,
       aes(x=joviality, fill= educationLevel)) +
  geom_histogram(bins=20, color='grey30')
```

```{r}
ggplot(data = participants_data,
       aes(x=age, fill= interestGroup)) +
  geom_histogram(bins=20, color='grey30')
  
```

```{r}
ggplot(data = participants_data,
       aes(x=joviality, fill= interestGroup)) +
  geom_histogram(bins=20, color='grey30')
  
```


```{r}
ggplot(data = participants_data,
       aes(x=age, fill= haveKids)) +
  geom_histogram(bins=20, color='grey30')
  
```

```{r}
ggplot(data = participants_data,
       aes(x=joviality, fill= haveKids)) +
  geom_histogram(bins=20, color='grey30')
```

```{r}
ggplot(data = participants_data,
       aes(y=age, x= educationLevel)) +
  geom_boxplot() +
  geom_point(stat='summary',
             fun.y='mean',
             colour='red',
             size=4)
```

```{r}
ggplot(data = participants_data,
       aes(y=age, x= haveKids)) +
  geom_boxplot() +
  geom_point(stat='summary',
             fun.y='mean',
             colour='red',
             size=4)
```



```{r}
ggplot(data = participants_data,
       aes(y=age, x= educationLevel)) +
  geom_violin(fill='light blue') +
  geom_boxplot(alpha=0.2) +
  geom_point(stat='summary',
             fun.y='mean',
             colour='red',
             size=4)
```

```{r}
ggplot(data = participants_data,
         aes(x=age, y= joviality)) +
  geom_point() +
  geom_smooth(size=0.5)
```

```{r}
ggplot(data = participants_data,
         aes(x=age)) +
  geom_histogram(bins=20) +
  facet_grid(~educationLevel)
```

```{r}
ggplot(data = participants_data,
         aes(y=age, x = interestGroup)) +
  geom_boxplot() +
  facet_grid(educationLevel~.)
```

```{r}
ggplot(data = participants_data,
         aes(y=joviality, x = interestGroup)) +
  geom_boxplot() +
  facet_grid(educationLevel~.)
```

```{r}
ggplot(data = participants_data,
         aes(y=age, x = interestGroup)) +
  geom_boxplot() +
  facet_grid(householdSize~.)
```

```{r}
ggplot(data = participants_data,
         aes(y=age, x = haveKids)) +
  geom_boxplot() +
  facet_grid(householdSize~educationLevel)
```

```{r}
ggplot(data = participants_data,
         aes(y=age, x = haveKids)) +
  geom_boxplot() +
  facet_grid(interestGroup~educationLevel) +
  coord_flip()
```




```{r}
ggplot(data = participants_data,
         aes(x=age, colour = educationLevel)) +
  geom_density()
```





